---
title: "Sleep"
output:
  html_document: default
  html_notebook: default
---

# Does brain mass predict how much mammals sleep in a day?

Original notebook by Tristan Mahr https://tjmahr.github.io/visualizing-uncertainty-rstanarm/.
Tristan's model had the problem that it would predict sleep times over 24h per day. 

Tristan's copyright: 

The MIT License (MIT)

Copyright (c) 2016 TJ Mahr

Permission is hereby granted, free of charge, to any person obtaining a copy
of this software and associated documentation files (the "Software"), to deal
in the Software without restriction, including without limitation the rights
to use, copy, modify, merge, publish, distribute, sublicense, and/or sell
copies of the Software, and to permit persons to whom the Software is
furnished to do so, subject to the following conditions:

The above copyright notice and this permission notice shall be included in all
copies or substantial portions of the Software.

THE SOFTWARE IS PROVIDED "AS IS", WITHOUT WARRANTY OF ANY KIND, EXPRESS OR
IMPLIED, INCLUDING BUT NOT LIMITED TO THE WARRANTIES OF MERCHANTABILITY,
FITNESS FOR A PARTICULAR PURPOSE AND NONINFRINGEMENT. IN NO EVENT SHALL THE
AUTHORS OR COPYRIGHT HOLDERS BE LIABLE FOR ANY CLAIM, DAMAGES OR OTHER
LIABILITY, WHETHER IN AN ACTION OF CONTRACT, TORT OR OTHERWISE, ARISING FROM,
OUT OF OR IN CONNECTION WITH THE SOFTWARE OR THE USE OR OTHER DEALINGS IN THE
SOFTWARE.

Let’s use the mammal sleep dataset from ggplot2. This dataset contains the number of hours spent sleeping per day for 83 different species of mammals along with each species’ brain mass (kg) and body mass (kg), among other measures. Here’s a first look at the data.
Preview sorted by brain/body ratio. The sorting was chosen so that humans show up in the preview.
```{r}
library(dplyr, warn.conflicts = FALSE)
library(ggplot2)
msleep %>% 
  select(name, sleep_total, brainwt, bodywt, everything()) %>% 
    arrange(desc(brainwt / bodywt))
```

Choose animals with known average brain weight, and add some transformed variables.
```{r}
msleep <- msleep %>% 
  filter(!is.na(brainwt)) %>% 
  mutate(log_brainwt = log10(brainwt), 
         log_bodywt = log10(bodywt), 
         log_sleep_total = log10(sleep_total),
         logit_sleep_ratio = qlogis(sleep_total/24))
```

Make a list of examples and give some familiar species shorter names
```{r}
ex_mammals <- c("Domestic cat", "Human", "Dog", "Cow", "Rabbit",
                "Big brown bat", "House mouse", "Horse", "Golden hamster")
renaming_rules <- c(
  "Domestic cat" = "Cat", 
  "Golden hamster" = "Hamster", 
  "House mouse" = "Mouse")

ex_points <- msleep %>% 
  filter(name %in% ex_mammals) %>% 
  mutate(name = stringr::str_replace_all(name, renaming_rules))
```

Define these labels only once for all the plots
```{r}
lab_lines <- list(
  brain_log = "Brain mass (kg., log-scaled)", 
  sleep_raw = "Sleep per day (hours)",
  sleep_log = "Sleep per day (log-hours)"
)
```

Plot sleep times vs. average brain weights
```{r}
ggplot(msleep) + 
  aes(x = brainwt, y = sleep_total) + 
  geom_point(color = "grey40") +
  # Circles around highlighted points + labels
  geom_point(size = 3, shape = 1, color = "grey40", data = ex_points) +
  ggrepel::geom_text_repel(aes(label = name), data = ex_points) + 
  # Use log scaling on x-axis
  scale_x_log10(breaks = c(.001, .01, .1, 1)) + 
    labs(x = lab_lines$brain_log, y = lab_lines$sleep_raw)
```

Next we use stan_glm from rstanarm package to make a linear model for logit of the sleep ratio given log of the brain weight (Tristan made the model with untransformed variables).
```{r}
library(rstanarm)
m1 <- stan_glm(
  logit_sleep_ratio ~ log_brainwt, 
  family = gaussian(), 
  data = msleep, 
  prior = normal(0, 3),
  prior_intercept = normal(0, 3))
```

Prepare x values for prediction:
```{r}
x_rng <- range(msleep$log_brainwt) 
x_steps <- seq(x_rng[1], x_rng[2], length.out = 80)
new_data <- data_frame(
  observation = seq_along(x_steps), 
  log_brainwt = x_steps)
```

Predict expected sleep time at new x values:
```{r}
preds<-posterior_linpred(m1,newdata=new_data)
preds<-plogis(preds)*24
```

Plot draws of the expected sleep time lines:
```{r}
library(reshape2)
gg<-data.frame(log_brainwt=new_data$log_brainwt,preds=t(preds[1:400,]))
gg<-melt(gg,id=c("log_brainwt"))
names(gg)<-c("log_brainwt","pp","preds")

# aesthetic controllers
alpha_level <- .15
col_draw <- "grey60"
col_median <-  "#3366FF"

ggplot(msleep) + 
  aes(x = log_brainwt, y = sleep_total) + 
  # Plot a random sample of rows as gray semi-transparent lines
  geom_line(aes(x=log_brainwt, y=preds, group=pp), 
              data = gg, color = col_draw, 
              alpha = alpha_level) + 
  geom_point() + 
  scale_x_continuous(labels = function(x) 10 ^ x) +
  labs(x = lab_lines$brain_log, y = lab_lines$sleep_raw)
```

Predict distribution of sleep times at new x values:
```{r}
preds_post <- posterior_predict(m1, newdata = new_data)
preds_post<-plogis(preds_post)*24
```

Plot distribution of sleep times at new x values:
```{r}
pq<-data.frame(t(apply(t(preds_post), 1, quantile, probs = c(0.025, 0.5, 0.995), na.rm = TRUE)))
names(pq)<-c("lower","median","upper")
pq$log_brainwt<-new_data$log_brainwt

ggplot(msleep) + 
  aes(x = log_brainwt) + 
  geom_ribbon(aes(ymin = lower, ymax = upper), data = pq, 
              alpha = 0.4, fill = "grey60") + 
  geom_line(aes(y = median), data = pq, colour = "#3366FF", size = 1) + 
  geom_point(aes(y = sleep_total)) + 
  scale_x_continuous(labels = function(x) 10 ^ x) +
      labs(x = lab_lines$brain_log, y = lab_lines$sleep_raw)
```
Here the distribution of the sleep times is restricted to be between 0 and 24 hours.
