---
title: "Bayesian data analysis - Student retention"
author: "Aki Vehtari"
date: "First version 2023-11-22. Last modified `r format(Sys.Date())`."
output:
  html_document:
    fig_caption: yes
    toc: TRUE
    toc_depth: 2
    number_sections: TRUE
    toc_float:
      smooth_scroll: FALSE
    theme: readable
    code_download: true
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(cache=FALSE, message=FALSE, error=FALSE, warning=TRUE, out.width='95%')
```

## Introduction

This includes the code used to create the models and plots for
student retention data used as an example in several BDA course
lectures in 2023.

#### Load packages

```{r}
library(brms)
#options(brms.backend = "cmdstanr")
# Using RStan backend for moment matching in LOO
options(brms.backend = "rstan")
library(posterior)
library(tidybayes)
library(dplyr)
library(tidyr)
library(ggplot2)
library(ggdist)
library(latex2exp)
library(khroma)
library(RColorBrewer)
theme_set(bayesplot::theme_default(base_family = "sans", base_size=16))
```

## Data

Since 2018, there have been 9 assignments in BDA course.  Data are
the number of students who submitted each assignment. As the course
is not compulsory for most, and it is common that some students
register for more courses than they need and during the course may
decide to drop-out. Also as the students can submit also in spring,
some students may stop submitting in autumn and get back in spring.
Although there are also external reasons why students drop out from
the course, we are interested in following the retention and hope
that the changes in the course would improve the retention. As a
starting point we want to analyse whether we can see differences in
years.


```{r}
# Number of students who returned the first assignment on 2018-2022
nstudents1<-rep(c(176,242,332,301,245),each=9)
# Number of students for each round
nstudents<-c(c(176, 174, 158, 135, 138, 129, 126, 123, 121),
             c(242, 212, 184, 177, 174, 172, 163, 156, 153),
             c(332, 310, 278, 258, 243, 242, 226, 224, 218),
             c(301, 269, 231, 232, 217, 208, 193, 191, 190),
             c(245, 240, 228, 217, 206, 199, 191, 182, 175))
# Proportion of students
propstudents<-c(c(176, 174, 158, 135, 138, 129, 126, 123, 121)/176,
                c(242, 212, 184, 177, 174, 172, 163, 156, 153)/242,
                c(332, 310, 278, 258, 243, 242, 226, 224, 218)/332,
                c(301, 269, 231, 232, 217, 208, 193, 191, 190)/301,
                c(245, 240, 228, 217, 206, 199, 191, 182, 175)/245)
# Year as integers and factors
year <- rep(2018:2022,each=9)
fyear <- factor(year)
# Assignment numbers
assignment <- rep(1:9, 5)
# Tibble
tb <- tibble(assignment, nstudents, nstudents1, propstudents, year, fyear)

# Another tibble for including 2023 first submission numbers
# Number of students who returned the first assignment
nstudents1<-rep(c(176,242,332,301,245,264),each=9)
# Number of students for each round
nstudents<-c(c(176, 174, 158, 135, 138, 129, 126, 123, 121),
                c(242, 212, 184, 177, 174, 172, 163, 156, 153),
                c(332, 310, 278, 258, 243, 242, 226, 224, 218),
                c(301, 269, 231, 232, 217, 208, 193, 191, 190),
                c(245, 240, 228, 217, 206, 199, 191, 182, 175),
                c(264,  NA,  NA,  NA,  NA,  NA,  NA,  NA,  NA))

propstudents<-c(c(176, 174, 158, 135, 138, 129, 126, 123, 121)/176,
                c(242, 212, 184, 177, 174, 172, 163, 156, 153)/242,
                c(332, 310, 278, 258, 243, 242, 226, 224, 218)/332,
                c(301, 269, 231, 232, 217, 208, 193, 191, 190)/301,
                c(245, 240, 228, 217, 206, 199, 191, 182, 175)/245,
                c(264,  NA,  NA,  NA,  NA,  NA,  NA,  NA,  NA)/264)
# Year as integers and factors
year <- rep(2018:2023,each=9)
fyear <- factor(year)
# Assignment numbers
assignment <- rep(1:9, 6)
# Tibble
tb2 <- tibble(assignment, nstudents, nstudents1, propstudents, year, fyear)

# Another tibble for including 2023 all submission rounds
# Number of students for each round
nstudents<-c(c(176, 174, 158, 135, 138, 129, 126, 123, 121),
                c(242, 212, 184, 177, 174, 172, 163, 156, 153),
                c(332, 310, 278, 258, 243, 242, 226, 224, 218),
                c(301, 269, 231, 232, 217, 208, 193, 191, 190),
                c(245, 240, 228, 217, 206, 199, 191, 182, 175),
                c(264, 249, 215, 221, 215, 206, 192,  186, 179))

propstudents<-c(c(176, 174, 158, 135, 138, 129, 126, 123, 121)/176,
                c(242, 212, 184, 177, 174, 172, 163, 156, 153)/242,
                c(332, 310, 278, 258, 243, 242, 226, 224, 218)/332,
                c(301, 269, 231, 232, 217, 208, 193, 191, 190)/301,
                c(245, 240, 228, 217, 206, 199, 191, 182, 175)/245,
                c(264, 249, 215, 221, 215, 206, 192, 186, 179)/264)
# Year as integers and factors
year <- rep(2018:2023,each=9)
fyear <- factor(year)
# Assignment numbers
assignment <- rep(1:9, 6)
# Tibble
tb3 <- tibble(assignment, nstudents, nstudents1, propstudents, year, fyear)
```

Plot of submission numbers

```{r}
ggplot(tb, aes(x=assignment, y=nstudents, group=year, color=as.factor(year-2017))) +
  geom_line(linewidth=1) + geom_point(size=3) +
  scale_x_continuous(breaks=1:9, lim=c(1,10.2)) +
  labs(x="Assignment", y="Number of students submitted", title="Student retention") +
  theme(legend.position="none") +
  scale_color_bright()+
  annotate(geom="text", x=rep(9.9, 6), y=c(121,153,218,192,169,180), label=c(2018:2023), size=5, color = color("bright")(6))
```

Plot of raw proportions *100 for different assignments and years

```{r}
ggplot(tb, aes(x=assignment, y=propstudents*100, group=year, color=as.factor(year-2017))) +
  geom_line(linewidth=1) + geom_point(size=3) +
  scale_x_continuous(breaks=1:9, lim=c(1,10.2)) +
  ylim(c(55,100)) +
  labs(x="Assignment", y="Proportion submitted %", title="Student retention") +
  theme(legend.position="none") +
  scale_color_bright()+
  annotate(geom="text", x=rep(9.9, 5), y=c(69, 61, 66, 63.5, 72), label=c(2018:2022), size=5, color = color("bright")(5))
```

## Models

Latent hierarchical linear model + binomial observation model

```{r}
# save_pars is used for later moment matching
fit4 <- brm(nstudents | trials(nstudents1) ~ assignment + (assignment | year), family=binomial(), data=filter(tb, assignment>1), control = list(adapt_delta = 0.95), save_pars=save_pars(all=TRUE), seed=7253)
```

First with plain PSIS-LOO, and we see some warnings

```{r}
fit4 <- add_criterion(fit4, 'loo', save_psis=TRUE, moment_match=FALSE, overwrite=TRUE)
loo(fit4)
```

PSIS-LOO + moment matching

```{r}
# overwrite is needed to force LOO recomputation
fit4 <- add_criterion(fit4, 'loo', save_psis=TRUE, moment_match=TRUE, overwrite=TRUE)
loo(fit4)
```

Latent spline + hierarchical linear model with binomial observation model

```{r}
fit6 <- brm(nstudents | trials(nstudents1) ~ s(assignment, k=4) + (assignment | year), family=binomial(), data=filter(tb, assignment>1), control = list(adapt_delta = 0.95), save_pars=save_pars(all=TRUE), seed=7253)
```

First with plain PSIS-LOO, and we see some warnings

```{r}
fit6 <- add_criterion(fit6, 'loo', save_psis=TRUE, moment_match=FALSE, overwrite=TRUE)
loo(fit6)
```

PSIS-LOO + moment matching

```{r}
# overwrite is needed to force LOO recomputation
fit6 <- add_criterion(fit6, 'loo', save_psis=TRUE, moment_match=TRUE, overwrite=TRUE)
loo(fit6)
```

Compare models

```{r}
loo_compare(loo(fit4), loo(fit6))
```

## Model predictions

Plot intervals for assignment 9 proportion estimates

```{r}
assign9linpred<-rvar(posterior_linpred(fit6, newdata=filter(tb,assignment==9), trandform=TRUE))
data.frame(year=c("2018","2019","2020","2021","2022"),propstudents=mean(assign9linpred),q05=quantile(assign9linpred,0.05)[1:5],q95=quantile(assign9linpred,0.95)[1:5])|>
ggplot(aes(x=year, y=propstudents*100, ymin=q05*100, ymax=q95*100)) +
  geom_pointrange(color=4) +
  labs(x="Year", y="Proportion submitted %", title="Proportion submitting 9th assign. (90% interval)")
```

Plot distribution of the difference in linear predictor

```{r}
assign9linpred<-rvar(posterior_linpred(fit6, newdata=filter(tb,assignment==9), transform=TRUE))
(assign9linpred[5]-assign9linpred[1:4]) |>
  as_draws_df() |>
  rename_with(~ c("2018","2019","2020","2021"), starts_with("x")) |>
  pivot_longer(cols=starts_with("2"), names_to="year", values_to="value") |>
  ggplot(aes(y=year, x=value)) +
  stat_halfeye() +
  geom_vline(xintercept=0) +
  ylab("Year")+
  xlab(TeX("2022 retention was better $\\rightarrow$"))
```

Plot model prediction of proportions for 2018-2022

```{r}
tb |>
  filter(assignment>1)|>
  add_linpred_draws(fit6, transform=TRUE) |>
  ggplot(aes(x = assignment, y = propstudents, group=fyear)) +
  stat_lineribbon(aes(y = .linpred), .width = c(.95), alpha = 1/2, color=brewer.pal(5, "Blues")[[5]])+
  geom_point(data=tb, color=1)+
  scale_fill_brewer() +
  facet_grid(. ~ fyear)+
  theme(legend.position="none")+
  labs(x="Assignment", y = "Proportion of submissions")+
  scale_x_continuous(breaks=1:9, lim=c(1,10.2))
```

Plot model prediction of proportions for 2018-2023 after we have
observed only the first assignment submission numbers for 2023

```{r}
tb2 |>
  filter(assignment>1)|>
  group_by(fyear) |>
  add_linpred_draws(fit6, transform=TRUE, allow_new_levels=TRUE) |>
  ggplot(aes(x = assignment, y = propstudents, group=fyear)) +
  stat_lineribbon(aes(y = .linpred), .width = c(.95), alpha = 1/2, color=brewer.pal(5, "Blues")[[5]])+
  geom_point(data=tb, color=1)+
  scale_fill_brewer() +
  facet_grid(. ~ fyear)+
  theme(legend.position="none")+
  labs(x="Assignment", y = "Proportion of submissions")+
  scale_x_continuous(breaks=1:9, lim=c(1,10.2))
```

Update the posterior with 2023 data

```{r}
fit6b <- update(fit6, newdata=filter(tb3, assignment>1))
```

Plot model prediction of proportions for 2018-2022 after observing
all submission numbers for 2023

```{r}
tb3 |>
  filter(assignment>1)|>
  add_linpred_draws(fit6b, transform=TRUE) |>
  ggplot(aes(x = assignment, y = propstudents, group=fyear)) +
  stat_lineribbon(aes(y = .linpred), .width = c(.95), alpha = 1/2, color=brewer.pal(5, "Blues")[[5]])+
  geom_point(data=tb3, color=1)+
  scale_fill_brewer() +
  facet_grid(. ~ fyear)+
  theme(legend.position="none")+
  labs(x="Assignment", y = "Proportion of submissions")+
  scale_x_continuous(breaks=1:9, lim=c(1,10.2))
```

The posterior predictive distribution for year 2023

```{r}
posterior_linpred(fit6, allow_new_levels=TRUE,
                  newdata=filter(tb3,year==2023&assignment==9),
                  transform = TRUE) |>
  quantile(c(0.05,0.95))
```

## PPC

PPC density overlays

```{r}
pp_check(fit4, ndraws=20)+
  xlim(c(50,370))
pp_check(fit6, ndraws=20)+
  xlim(c(50,370))
```

PPC intervals grouped

```{r}
pp_check(fit4, type = "intervals_grouped", group="year",
         facet_args=list(nrow=1))
pp_check(fit6, type = "intervals_grouped", group="year",
         facet_args=list(nrow=1))
```

PPC ribbon grouped

```{r}
pp_check(fit4, type = "ribbon_grouped", group="year",
         facet_args=list(nrow=1))
pp_check(fit6, type = "ribbon_grouped", group="year",
         facet_args=list(nrow=1))
```

PPC intervals

```{r}
pp_check(fit4, type = "intervals")
pp_check(fit6, type = "intervals")
```

PPC LOO intervals

```{r}
pp_check(fit4, type = "loo_intervals")
pp_check(fit6, type = "loo_intervals")
```

PPC LOO-PIT-QQ-uniform

```{r}
pp_check(fit4, type = "loo_pit_qq", ndraws=4000)
pp_check(fit6, type = "loo_pit_qq", ndraws=4000)
```

PPC LOO-PIT-QQ-normal

```{r}
pp_check(fit4, type = "loo_pit_qq", ndraws=4000, compare="normal")
pp_check(fit6, type = "loo_pit_qq", ndraws=4000, compare="normal")
```

